---
title: "mecRecordLinkage"
author: "Łukasz Chrostowski"
output:
  html_document: default
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
editor: visual
execute: 
  eval: true
  warning: false
  message: false
library:
  - dplyr
  - reclin2
---

## Introduction

The goal of the main function `mec` available in the package is to link these sets using maximum entropy classifier. Combining different sources of data is widely using step in order to make an data analysis. Many various techniques have been considering during the years so as to link the records base on a comparison vectors and calculates m- and u- probabilities. The most useful is probabilistic record linkage derived by `Fellegi and Sunter`, where parameters are estimated using `EM` algorithm (`Winkler`) (see `reclin2` package). One can use also machine learning methods, where problem of matching pairs of records into match and non-match sets is considering as a classification problem. Using `glm` with `binomial` or `xgboost` are one of the proposed methods. Here pairs matching is a classification problem as well, but matching process is more automated in comparison to probabilistic method. MEC technique was proposed by D. Lee, L. C. Zhang and J. K. Kim "Maximum entropy classification for record linkage" (2022). Authors derive supervised and unsupervised methods with full set parameters estimation techniques. `mecRecordLinkage` package contains direct implementation of the proposed algorithms. First of all record linkage using `mec` depending on the estimation techniques will be performed. Then we compare results with the `reclin2` package. In contrast to `reclin2` package, there is one function to prepare data sets, create comparison vectors using `reclin2` package, obtain parameters and link record.

## Technical details

With record linkage we try to link two or more data sets using many common entities. The goal is to find the true matches among all possible pairs of the two data files. For this purposes we create so called `comparison space` $\Omega=A \times B=M \cup U$ consist of matches M and non-matches U between the record in given files. First of all we want to estimate probability ratio

$$
r_{a b}=\frac{m\left(\gamma\right)}{u\left(\gamma\right)}
$$ where $m\left(\gamma\right)$ and $u\left(\gamma\right)$ are the probability mass functions for M and U set directly. In maximum entropy classification we classifiy each pair of records to these sets depending of the $r$ value. We consider several ways to estimate probability ratio. Let $\quad g_{a b}=1$ if $\quad(a, b) \in M \quad$ and $0$ if $\quad(a, b) \in U$. In the first setting we can create a model of $m\left(\gamma\right)$ by $m(\gamma ; \boldsymbol{\theta})=\prod_{k=1}^K \theta_k^{\gamma_k}\left(1-\theta_k\right)^{1-\gamma_k}$. where $\theta_k=\operatorname{Pr}\left(\gamma_{a b, k}=1 \mid g_{a b}=1\right)$, and $\gamma_{a b, k}$ is the $k^{\text {th }}$ component of $\gamma_{a b}$. $u(\gamma ; \xi)$ can be modeled as above with parameters $\xi_k$ instead of $\theta_k$, where $\xi_k=\operatorname{Pr}\left(\gamma_{a b, k}=1 \mid g_{a b}=0\right)$. The goal in the MEC method is to find the set of pairs that maximize the entropy function. There is different possibilities for estimating required parameters. One the one hand we can use number of agreements (non-agreements) on the key variables, on the other hand maximum likelihood method is considering as well. In the default setting we have

$$\theta_k^{(t)}=\frac{1}{\left|M^{(t)}\right|} \sum_{(a, b) \in \Omega} g_{a b}^{(t)} \mathbb{I}\left(\gamma_{a b, k}=1\right)$$

Another propose is to derive $\theta_k$ from all the pairs in $\Omega$ whereas $\theta_k$ given as above uses only the pairs from the MEC set $M$, then we have

$$\theta_k^{(t)}=\frac{1}{n_M^{(t)}} \sum_{(a, b) \in \Omega} \hat{g}_{a b}^{(t)} \gamma_{a b, k}$$

For $\xi_k$ estimation we consider rate agreements from all the pairs in $\Omega$ in which case there is no updating of $u\left(\gamma ; \xi^{(t)}\right)$ in the iterative algorithm for the unsupervised setting.

$$\hat{\xi}_k=\frac{1}{n} \sum_{(a, b) \in \Omega} \mathbb{I}\left(\gamma_{(a b, k)}=1\right)$$

MEC for record linkage should generally be guided by the error rates, false link rate (FLR) and missing match rate (MMR), for whose bisection procedure by a threshold value is derived. Errors are defining as follows

$$\psi=\frac{1}{|\hat{M}|} \sum_{(a, b) \in \hat{M}}\left(1-g_{a b}\right)$$

$$\tau=1-\frac{1}{n_M} \sum_{(a, b) \in \hat{M}} g_{a b}$$

for FLR and MMR respectively.

## MEC for unsupervised learning - iterative algorithm

One can divide maximum entropy classification into two machine learning techniques - supervised and unsupervised learning. The second method provides iterative algorithm to obtain terminal result. D. Lee, L. C. Zhang and J. K. Kim propose following procedure.

1.  set $\boldsymbol{\theta}^{(0)}=\left(\theta_1^{(0)}, \ldots, \theta_K^{(0)}\right)$ and $n_M^{(0)}=\left|M_1\right|$ , where $M_1$ is the maximal MEC set (only consists of the record pairs with perfect agreement of all the key variables).
2.  For the i-th iteration let $g_{a b}^{(t)}=1$ if $(a, b) \in M^{(t)}$, and 0 otherwise.

```{=html}
<!-- -->
```
i)  update $u\left(\gamma ; \xi^{(t)}\right)$ and $m(\gamma ; \theta^{(t)})$, given $\mathbf{g}^{(t)}=\left\{g_{a b}^{(t)}:(a, b) \in \Omega\right\}$, using one of the estimation methods described above. Once $\boldsymbol{\theta}^{(t)}$ and $\xi^{(t)}$ are obtained, we can update $n_M^{(t)}=\sum_\gamma n(\gamma) \hat{g}^{(t)}(\gamma)$, where $$
    \hat{g}^{(t)}(\gamma) \equiv \hat{g}\left(\gamma ; \boldsymbol{\theta}^{(t)}, \xi^{(t)}\right)=\min \left\{\frac{\left|M^{(t)}\right| r^{(t)}(\gamma)}{\left|M^{(t)}\right|\left(r^{(t)}(\gamma)-1\right)+n}, 1\right\}
    $$ and $r^{(t)}(\gamma)$ is a probability ratio for current setting.
ii) For given $n_{M}^{(t+1)}$ we find the maximal entropy classification set $M^{(t+1)} = =\left\{(a, b) \in \Omega: g_{a b}^{(t+1)}=1\right\}$ such that $\left|M^{(t+1)}\right|=n_M^{(t)}$ by deduplication in the descending order of probablity ratio over $\Omega$. It maximizes the entropy function.

```{=html}
<!-- -->
```
3.  Iterate until $n_M^{(t)}=n_M^{(t+1)}$ or $\left\|\boldsymbol{\theta}^{(t)}-\boldsymbol{\theta}^{(t+1)}\right\|<\epsilon$.

## Package installation and loading

In order to use `mecRecordLinkage` package effectively there is no need to install or load additional packages. All functionalities are build in. To prepare datasets properly using`reclin2` packages is proposed inside of the `mecRecordLinkage`.

```{r setup, echo=FALSE}
devtools::load_all()
#remotes::install_github("ncn-foreigners/mecRecordLinkage")
#library(mecRecordLinkage)
```

## Work with example datasets and initial exploration

To perform example of using `mec` function we can use following data sets. Data are obtained from https://github.com/djvanderlaan/tutorial-reclin-uros2021 and contain columns with the most important attributes like first name, last name or sex related to a given person (for more see documentation).

```{r}
data1 <- read.csv("data/example1.csv", stringsAsFactors = FALSE)
data2 <- read.csv("data/example2.csv", stringsAsFactors = FALSE)
head(data1)
head(data2)
```

## Unsupervised learning using `mec` function

`mec` Function enables to use unsupervised learning. In the first case we link data sets on given variables assuming that year of birthday given person is strata.

```{r}
vars <- c("firstname", "lastname", "sex", "dobday", "dobmon")
model <- mec(A = data1, B = data2, vars = vars, blockvars = "dobyear")
```

Let's look at the return of the model.

Parameters of the model

```{r}
model$theta # for m-probability
model$eta # for u-probability
model$matching_prob # Expected value of being in the M set
model$class_entropy # value of the entropy function for the final result
```

Model returns M set, its size and linked datasets as well.

```{r}
model$M
```

Depending on the estimation method, results can be different. You can define that by yourself with the `control` parameter.

```{r}
model2 <- mec(A = data1, B = data2, vars = vars, blockvars = "dobyear", control = control_mec(theta_est = "2"))
model2$M
```

## Guiding the algorithm by the error rates

```{r}
model_ER1 <- mec(A = data1, B = data2, vars = vars, blockvars = "dobyear", error_rate = TRUE)
model_ER1$M
```

```{r}
#model_ER2 <- mec(A = data1, B = data2, vars = vars, blockvars = "dobyear", control = control_mec(theta_est = "2"), error_rate = TRUE)
#model_ER2$M
```

## Other packages for record linkage

Let's compare it to the results for `reclin2` package, where `EM` algorithm is implemented.

```{r}
pairs <- reclin2::pair_blocking(data1, data2, on = "dobyear")
reclin2::compare_pairs(pairs, on = vars, inplace = TRUE)
m <- reclin2::problink_em(~lastname + firstname + 
                            sex + dobday + dobmon, data = pairs)

pairs <- predict(m, pairs = pairs, add = TRUE)

pairs <- reclin2::select_threshold(pairs, "threshold", score = "weights", 
                                   threshold = 8)
pairs <- reclin2::select_n_to_m(pairs, "weights", variable = "ntom", 
                                threshold = 0)

linked_data <- reclin2::link(pairs, selection = "ntom", all = FALSE)
  
```

Notice that this method requires defining threshold by the user, while in maximal MEC there is no need to define that, what makes this method more automated.

## Supervised learning
